A note on observation processes in epidemic models

### 2019-07-30 ms/ms.tex
keep tweaking

Epidemic models provide valuable tools for understanding how a disease
spreads in a population. Many epidemic models focus on characterizing
the underlying transmission mechanism but make naive assumptions about
how infections are reported. In this note, we use a simple deterministic
Susceptible-Infected-Removed (SIR) model to compare two popular assumptions 
about disease incidence reports: individuals can report their infection as soon as
they become infected or as soon as they recover. We show that making
wrong assumptions about the underlying observation processes can result
in biased estimates of the basic reproduction number and narrow confidence
intervals.

@@

Mechanistic analyses of epidemic time series allow us to infer the underlying 
transmission mechanism, estimate biologically relevant parameters, and 
predict the course of an outbreak \citep{breto2009time}. In order to make a 
precise and accurate inference, disease modelers try to build
more realistic process models. For example, the same London measles time 
series from the prevaccination era has been analyzed using multiple models -- 
these models account for time-varying transmission rates \citep{fine1982measles}, 
realistic age structure \citep{schenzle1984age},
metapopulation structure \citep{xia2004measles}, continuous-time infection process 
\citep{cauchemez2008likelihood}, and extra-demographic variability \citep{he2009plug}.

Despite the amount of effort put into developing \emph{better} process models, disease 
modelers often neglect details of the observation process associated with disease
case reports. Most disease models effectively assume that new cases are reported
instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
kennedy2018modeling, pons2018serotype}) or develops symptom (e.g., 
\cite{bhadra2011malaria, king2015avoidable}); 
a few models (e.g.,\cite{breto2009time, he2009plug, lin2016seasonality})
assume that infections are counted upon recovery (because diagnosed cases
are controlled and are effectively no longer infectious).

@@

For brevity, we consider two extreme cases: individuals instantaneously report
their infection when they become infected or when they recover. The observed incidence 
measured upon infection, $i_1(t)$, can be defined by the integral:
\begin{equation}
i_1(t) = \int_{t - \Delta t}^{t} \beta S \frac{I}{N} dt.
\end{equation}
Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
new state variable described by $dC/dt = \beta S I/N$ and taking the difference between 
the two consecutive reporting periods: $i_1(t) = C(t) - C(t-\Delta t)$. Likewise, 
incidence measured upon recovery, $i_2(t)$, can be defined by the integral:
\begin{equation}
i_2(t) = \int_{t-\Delta t}^{t} \gamma I dt,
\end{equation}
or by the consecutive difference in the cumulative number of recovered cases:
$i_2(t) = R(t) - R(t - \Delta t)$.
Finally, we model under-reporting using a negative binomial distribution with a
mean either $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
an over-dispersion parameter $\theta$. For convenience, we will refer to these two
negative binomial models as infection model and recovery model hereafter; 
likewise, we will refer to epidemic time series generated from these
negative binomial models 
as infection time series and recovery time series.

@@

First, we compare the deterministic dynamics of incidence curves, 
$i_1(t)$ and $i_2(t)$, 
(Figure 1). A lag in reporting time delays
the observed epidemic peak timing and reduces the size of the peak. However, 
these differences in the reporting time have small effect to the overall shape 
of the epidemic curve. In the presence of observation and process error, we 
might not expect to be able to distinguish between the two reporting processes
based on the time series alone.

@@

In order to understand how assumptions about the timing of case reporting affect 
parameter estimates of the SIR model, we simulate observed epidemic curves 
(i.e., infection time series and recovery time series), measured upon both infection and
recovery, 100 times and fit both models (i.e., infection model and recovery model) to each 
time series. We compare the estimates of the basic reproductive number $\mathcal R_0$,
and its coverage probability, defined as the proportion of confidence intervals that
contain the true value (95\% confidence interval is expected to contain the true value
95\% of the time by definition). We summarize the results in Figure 2.

@@

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (e.g.,
\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
When we assume that the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model, fitting the
wrong model results in a clearer bias and a much lower coverage (Figure 2B); however, 
bias from fitting the wrong model changes direction (e.g., fitting the infection
model to recovery model overestimates the reproduction number instead).

@@

While we only considered observation processes associated with incidence 
reporting in this study, it is equally as important to consider observation
processes in mortality reporting. In particular, many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that individual deaths are recorded as soon as individuals die 
\citep{he2013inferring, didelot2017model, dean2018human}; this includes
\cite{kermack1927contribution} who approximated the reported \emph{number} of
deaths per week from plague with instantaneous death \emph{rates} ($dR/dt$).
This assumption does not account for the possibility 
that a delay in reporting of deaths can
change the shape of an epidemic curve: longer delays are expected to decrease
the size of an epidemic peak and delay the timing of the peak (Figure 1). 
Ignoring delays in reporting may affect their conclusions about the underlying 
transmission mechanisms.

>>>>

Epidemic models provide valuable tools for understanding how a disease
spreads in a population. Many epidemic models focus on characterizing
the underlying transmission mechanism but make naive assumptions about
how infections are reported. In this note, we use a simple deterministic
Susceptible-Infected-Removed (SIR) model to compare two common assumptions 
about disease incidence reports: individuals can report their infection as soon as
they become infected or as soon as they recover. We show that making
wrong assumptions about the underlying observation processes can result
in biased estimates of the basic reproduction number and narrow confidence
intervals.

@@

Mechanistic analyses of epidemic time series allow us to infer the underlying 
transmission mechanism, estimate biologically relevant parameters, and 
predict the course of an outbreak \citep{breto2009time}. In order to make a 
precise and accurate inference, disease modelers have sought to build
more realistic process models. For example, the same London measles time 
series from the prevaccination era has been analyzed many times -- 
these models account for time-varying transmission rates \citep{fine1982measles}, 
realistic age structure \citep{schenzle1984age},
metapopulation structure \citep{xia2004measles}, continuous-time infection process 
\citep{cauchemez2008likelihood}, and extra-demographic variability \citep{he2009plug}.

Despite the amount of effort put into developing \emph{better} process models, disease 
modelers often neglect details of the observation process associated with disease
case reports. Many disease models effectively assume that new cases are reported
instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
kennedy2018modeling, pons2018serotype}) or develops symptom (e.g., 
\cite{bhadra2011malaria, king2015avoidable}); 
some models (e.g., \cite{breto2009time, he2009plug, lin2016seasonality})
assume that infections are counted upon recovery (because diagnosed cases
are controlled and are effectively no longer infectious).

@@

For brevity, we consider two extreme cases: individuals instantaneously report
their infection when they become infected or when they recover. The observed incidence 
measured upon infection, $i_1(t)$, can be defined by the integral:
\begin{equation}
i_1(t) = \int_{t - \Delta t}^{t} \beta S \frac{I}{N} dt.
\end{equation}
Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
new state variable described by $dC/dt = \beta S I/N$ and taking the difference between 
the two consecutive reporting periods: $i_1(t) = C(t) - C(t-\Delta t)$. Likewise, 
incidence measured upon recovery, $i_2(t)$, can be defined by the integral:
\begin{equation}
i_2(t) = \int_{t-\Delta t}^{t} \gamma I dt,
\end{equation}
or by the consecutive difference in the cumulative number of recovered cases:
$i_2(t) = R(t) - R(t - \Delta t)$.
Finally, we model under-reporting using a negative binomial distribution with a
mean either $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
an over-dispersion parameter $\theta$. For convenience, we will refer to these two
negative binomial models as infection model and recovery model hereafter; 
likewise, we will refer to epidemic time series generated from these
negative binomial models with two different means ($\rho i_1(t)$ and $\rho i_2(t)$)
as infection time series and recovery time series.

@@

First, we compare the deterministic dynamics of incidence curves,
$i_1(t)$ and $i_2(t)$, in Figure 1. A lag in reporting time delays
the observed epidemic peak timing and reduces the size of the peak. However, 
these differences in the reporting time have small effect to the overall shape 
of the epidemic curve. In the presence of observation and process error, we 
might not expect to be able to distinguish between the two reporting processes
based on the time series alone.

@@

In order to understand how assumptions about the timing of case reporting affect 
parameter estimates of the SIR model, we simulate observed epidemic time series
(infection time series and recovery time series) 
100 times and fit both infection and recovery models to each 
time series. We compare the estimates of the basic reproductive number $\mathcal R_0$,
and its coverage probability, defined as the proportion of confidence intervals that
contain the true value (95\% confidence interval is expected to contain the true value
95\% of the time by definition). We summarize the results in Figure 2.

@@

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (e.g.,
\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
When we assume that the true value of the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model, fitting the
wrong model results in a clearer bias and a much lower coverage (Figure 2B); however, 
bias changes direction (e.g., fitting the infection
model to recovery model overestimates the reproduction number instead).

@@

While we only considered observation processes associated with incidence 
reporting in this study, it is equally as important to consider observation
processes in mortality reporting. In particular, many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that individual deaths are recorded as soon as individuals die 
\citep{he2013inferring, didelot2017model, dean2018human}; this includes
\cite{kermack1927contribution} who approximated the reported \emph{number} of
deaths per week from plague with instantaneous death \emph{rates} ($dR/dt$).
This assumption does not account for the possibility 
that a delay in reporting of deaths can
change the shape of an epidemic curve: delayed reporting is expected to decrease
the size of the observed epidemic peak and delay the observed timing of the peak (Figure 1). 
Ignoring delays in reporting may affect conclusions about the underlying 
transmission mechanisms.
